---
title: "Problem Set 4"
subtitle: "Political Data Science - Spring 2020"
date: "Due May 1, 10:00 AM (Before Class)"
author: "Gencer, Alper Sukru"
output: pdf_document
header-includes:
  - \usepackage{color}
  - \usepackage{float}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, fig.pos = 'H')
```






\section{Instructions}

\bigskip

\begin{enumerate}
  \item The following questions should each be answered within an R script. Be sure to provide many comments
in the script to facilitate grading. Undocumented code will not be graded.
  \item Work on git. Fork the repository found at https://github.com/domlockett/PDS-PS3 and add your
code, committing and pushing frequently. Use meaningful commit messages – these may affect your
grade.
  \item You may work in teams, but each student should develop their own R script. To be clear, there should
be no copy and paste. Each keystroke in the assignment should be your own.
  \item If you have any questions regarding the Problem Set, contact the TAs or use their office hours.
  \item For students new to programming, this may take a while. Get started.
  \item You will need to install ggplot2 and dplyr to complete this dataset.
\end{enumerate}






\newpage
\section{Question 1 - Sample Statistics}

\bigskip

Load the following data: http://politicaldatascience.com/PDS/Datasets/GSS-data.csv. 

\bigskip

The variable poleff11 asks participants to rate their level of agreement with the statement “People like me don’t have any say about what the government does” (see the codebook for more information on all variables in this dataset at: http://politicaldatascience.com/PDS/Datasets/gss_codebook.csv).

\bigskip

\begin{enumerate}
  \item Convert this variable into a numeric where higher values indicate higher levels of political efficacy (1- strongly agrees with the statement; 5- strongly disagrees with the statment) and all other values (‘Cant choose’ etc.) become NA’s. 
  \item What is the proportion of individuals from the entire sample who feel as though they "have a say in the government?"
  \item Using a sample of 25 from this dataset. What is the average proportion who feel as thought hey have a say?
  \item Pull a random sample of 25 from the poleff11 data and calculate the mean for this outcome. Now repeat this process 500 times and store these values in a variable called trials 25.
  \item Now create a variable called trials 100 where we do 500 trials with $n = 100$ instead of $25$.
  \item Draw a histogram of the sampling distribution for the two trials $(n=25$ vs. $n=100)$ you just conducted. Give the plots meaningful titles and axis labels. Save these plots in your repository.
  \item What notable difference occur when we use a larger sample size in our trials?
\end{enumerate}

\newpage
\section{Answer 1 - Sample Statistics}

Load the following data: http://politicaldatascience.com/PDS/Datasets/GSS-data.csv. 

\bigskip

```{r getting ready Q1, include=TRUE, warning = FALSE, message = FALSE}
rm(list = ls())
gss.data <- read.csv("http://politicaldatascience.com/PDS/Datasets/GSS-data.csv")
gss.data <- gss.data[-c(2349, 2350), ] ## included some unrelated information
```

\bigskip

\subsection{Convert this variable into a numeric where higher values indicate higher levels of political efficacy (1- strongly agrees with the statement; 5- strongly disagrees with the statment) and all other values (‘Cant choose’ etc.) become NA’s:}

```{r Q1 Part 1, include=TRUE, warning = FALSE, message = FALSE}
levels(gss.data$poleff11)
gss.data$poleff11.recoded[gss.data$poleff11 == "Strongly agree"] <- 1
gss.data$poleff11.recoded[gss.data$poleff11 == "Agree"] <- 2
gss.data$poleff11.recoded[gss.data$poleff11 == "Neither agree nor disagree"] <- 3
gss.data$poleff11.recoded[gss.data$poleff11 == "Disagree"] <- 4
gss.data$poleff11.recoded[gss.data$poleff11 == "Strongly disagree"] <- 5
gss.data$poleff11.recoded[gss.data$poleff11 == "Cant choose"] <- NA
gss.data$poleff11.recoded[gss.data$poleff11 == "Not applicable"] <- NA
gss.data$poleff11.recoded[gss.data$poleff11 == "No answer"] <- NA
```

\bigskip

\subsection{What is the proportion of individuals from the entire sample who feel as though they "have a say in the government?":}

```{r Q1 Part 2, include=TRUE, warning = FALSE, message = FALSE}

########  For this, I'll be looking at the proportion of people who "disagree (4) \ strongly disagree (5):" 

#---  including NA values:
prop_have.a.say.na <-sum(table(gss.data$poleff11.recoded)[4:5])/nrow((gss.data))
prop_have.a.say.na

#---  excluding NA values:
prop_have.a.say <-sum(table(gss.data$poleff11.recoded)[4:5])/length(na.omit(gss.data$poleff11.recoded))
prop_have.a.say

```

\bigskip

\subsection{Using a sample of 25 from this dataset. What is the average proportion who feel as thought hey have a say?:}

```{r Q1 Part 3, include=TRUE, warning = FALSE, message = FALSE}
set.seed(20200425)

#---  including NA values:
sample25_na <-sample(gss.data$poleff11.recoded, size =25)
sum(sample25_na>=4, na.rm = T)/25

#---  excluding NA values:
sample25_no.na <-sample(na.omit(gss.data$poleff11.recoded), size =25)
sum(sample25_no.na>=4, na.rm = T)/25
```


\bigskip

\subsection{Pull a random sample of 25 from the poleff11 data and calculate the mean for this outcome. Now repeat this process 500 times and store these values in a variable called \textbf{trials_25}:}

```{r Q1 Part 4, include=TRUE, warning = FALSE, message = FALSE}

########  For convenience, I will be solving by removing NA's. 

#---  Random one sample:
mean(sample(na.omit(gss.data$poleff11.recoded), size =25, replace = T))


#---  Random 500 samples:
trials_25 <- NULL
for(i in 1:500){
  trials_25 <- c(trials_25, mean(sample(na.omit(gss.data$poleff11.recoded), size =25, replace = T)))
}

```



\bigskip

\subsection{Now create a variable called trials_100 where we do 500 trials with n=100 instead of 25:}

```{r Q1 Part 5, include=TRUE, warning = FALSE, message = FALSE}

########  For convenience, I will be solving by removing NA's. 

#---  Random one sample with n = 100:
mean(sample(na.omit(gss.data$poleff11.recoded), size =100, replace = T))


#---  Random 500 samples with n = 100:
trials_100 <- NULL
for(i in 1:500){
  trials_100 <- c(trials_100, mean(sample(na.omit(gss.data$poleff11.recoded), size =100, replace = T)))
}

```



\bigskip

\subsection{Draw a histogram of the sampling distribution for the two trials (n=25 vs. n=100) you just conducted. Give the plots meaningful titles and axis labels. Save these plots in your repository.:}

```{r Q1 Part 6, include=TRUE, warning = FALSE, message = FALSE, fig.align = "center", fig.cap = "Sampling Distributions with Varying Sample Sizes"}
library(tidyverse)
##install.packages("ggpubr")
library(ggpubr)

df.trials <- as.data.frame(cbind(trials_25, trials_100))

p_trials_100 <- ggplot(df.trials) +
  geom_histogram(aes(trials_100), bins = 100, fill = "red") +
  xlab("Sample Means") +
  ggtitle("Sampling Dist. with Size 100") +
  ylim(0,50) +
  xlim(2,4) +
  theme_light()

p_trials_25 <- ggplot(df.trials) +
  geom_histogram(aes(trials_25), bins = 100, fill = "blue") +
  xlab("Sample Means") +
  ggtitle("Sampling Dist. with Size 25") +
  ylim(0,50) +
  xlim(2,4) +
  theme_light()

ggarrange(p_trials_100, p_trials_25,
                    ncol = 2, nrow = 1)

```


\bigskip

\subsection{What notable difference occur when we use a larger sample size in our trials?:}

\bigskip

As it can be seen in the graph, the spead of the sampling distribution with sample size 100 is narrower than the one with sample size 25.

```{r Q1 Part 7, include=TRUE, warning = FALSE, message = FALSE}
library(knitr)

table <- cbind(mean(na.omit(gss.data$poleff11.recoded)), mean(trials_100), 
      sd(trials_100), mean(trials_25), sd(trials_25))

colnames(table) <- c("True Mean", "Size-100 Mean", "Size-100 SD", 
                  "Size-25 Mean", "Size-25 SD")

kable(table)
```

\bigskip

Also see from the table that the mean of size-100 trials is closer to the true mean and at the same time, the sampling distribution of sample mean has less variance (smaller standard error).

\newpage
\section{Q2 - tidyverse}